---
title: "School Enrollment Project"
subtitle: "Stat Analysis"
format: 
  html:
    toc: true
    toc_float: true 
editor: visual
execute:
  echo: true
---

# Load Packages 
```{r}
library(readr)
library(tidyverse)
```

# Import Data
```{r}
schools <- read_csv("schools.csv")

glimpse(schools)
```
# Statistical Tests 
## 1. Techniques like ARIMA (AutoRegressive Integrated Moving Average) could be useful if we are forecasting future trends based on past data. 
```{r}

```

## 2.To understand the differential impact of Covid-19 on various demographic groups, comparative analyses like t-tests (for comparing two groups) or ANOVA (for more than two groups) can be employed.       
```{r}

```

## 3.To examine the relationships between different variables without assuming causality, correlation analysis can be helpful.

```{r}

```

## 4. Regression Analysis: To explore relationships between variables (e.g., how economic status or race affects enrollment changes during Covid-19).
```{r}

```

